Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis

### 2019-12-18 article/Old versions/JAS/PCADSC_article_JAS_v1.tex
old changes to meta files

\articletype{Application Note}% Specify the article type or omit as appropriate

@@

PCADSC compares the covariance matrices of two subsets of a dataset. If all variables in the subsets are jointly normal with known means, the covariance matrices are sufficient statistics describing the joint distribution of all the variables, but even without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities describing linear interrelations between variables. This makes the empirical covariance matrix a reasonable place to start looking for differences in data structures. Computing empirical covariance matrices for the two subsets and comparing them entry-by-entry becomes increasingly difficult as the number of variables increases and for this reason we decompose and recompose the covariance matrices using \textit{principal component analysis} (PCA) to get an overview of differences between the two subsets. %We refer to Appendix \ref{appendix.introPCA} for a brief introduction to PCA.

@@

In \textit{model 1} the observed variables $V_3$ and $V_4$ are caused by a common, unobserved variable $L_1$. Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models, in Appendix \ref{appendix.simData}. In dataset $\mathbf{A}$, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures. In dataset $\mathbf{B}$, on the other hand, there are two fundamentally different, underlying data structures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.

@@

>>>>

\articletype{Original Research Article}% Specify the article type or omit as appropriate

@@

PCADSC compares the covariance matrices of two subsets of a dataset. If all variables in the subsets are jointly normal with known means, the covariance matrices are sufficient statistics describing the joint distribution of all the variables, but even without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities describing linear interrelations between variables. This makes the empirical covariance matrix a reasonable place to start looking for differences in data structures. Computing empirical covariance matrices for the two subsets and comparing them entry-by-entry becomes increasingly difficult as the number of variables increases and for this reason we decompose and recompose the covariance matrices using \textit{principal component analysis} (PCA) to get an overview of differences between the two subsets. We refer to Appendix \ref{appendix.introPCA} for a brief introduction to PCA.

@@

In \textit{model 1} the observed variables $V_3$ and $V_4$ are caused by a common, unobserved variable $L_1$. Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models, in Appendix \ref{appendix.introPCA}. In dataset $\mathbf{A}$, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures. In dataset $\mathbf{B}$, on the other hand, there are two fundamentally different, underlying data structures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.

@@

\section{A brief introduction to principal component analysis}
\label{appendix.introPCA}

In this appendix we introduce PCA. We suppress the subscripts and superscripts used in the article because we only consider a single dataset here, denoted by $\mathbf{X} \in \RR^{d \times n}$.

For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost. More specifically, we wish to minimize the loss when looking at a projection of $\mathbf{X}$ onto a $q$-dimensional space, rather than the original $d$-dimensional one. Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is

\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_{i \cdot} - \bar{X} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_{i \cdot} - \bar{X} - \text{proj}_{K_q}(x_{i \cdot} - \bar{X}) \rVert^2.
\end{equation*}

The theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error \cite{HastieEtAl2009}. More specifically, the rank-q-reconstruction error is attained when we choose

\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\},
\end{equation*}
where $\eta_1, ..., \eta_q$ are the first $q$ eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, \dotsc, \lambda_d$. In the PCA framework, we refer to the eigenvectors as \textit{loadings}. The eigenvalues may be understood as \textit{variance components}, as the sum of the marginal empirical variances is preserved under eigenvalue decomposition, that is,
\begin{equation*}
\text{trace}(S) = \sum_{j=1}^d \hat{V}(X_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top \mathbf{X}^\top) = \sum_{j=1}^d \lambda_j,
\end{equation*}
where $X_j \in \RR^n$ denotes the $j$th variable of $\mathbf{X}$ and $\hat{V}$ is the empirical variance function. This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe. And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are distinct, the representation hereby obtained is a valid object of inference.

The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$ that maximizes the variation of the associated scores:
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_{i \cdot} - \bar{X}) \rVert^2, &
\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_{i \cdot} - \bar{X}) \rVert^2,
\end{align*}
where the initial subspace is defined as $\hat{K}_0 = \{0\}$. It is worth emphasizing that this greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error.

At this point a general remark about standardization is in place. PCA deconstructs the covariance matrix in components according to the most explained variance. This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will always be deemed highly influential, no matter the structure of the data. Therefore, the variables are typically standardized prior to performing PCA. The covariance matrix of the standardized variables is the same as the correlation matrix of the original variables, so post-standardization PCA simply corresponds to performing data structure comparisons of the correlation matrices rather than the covariance matrices. The standardization makes the variables comparable on the same scale, namely units of standard deviation. We have adopted this standardization policy.
